Neural Connectome of the Ctenophore Statocyst

### 2025-06-16 manuscript/Manuscript.qmd
MS: edits based on comments

For vEM analysis, we used a five-day-old cydippid larva of *Mnemiopsis leidyi*. To obtain high-quality ultrastructural preservation, we used high pressure freezing followed by freeze substitution and resin embedding (Epon). The specimen was sectioned from the aboral tip and we collected approximately 1,000 ultra-thin (50 nm) serial sections as ribbons. The sections were imaged in a Zeiss Gemini 500 scanning electron microscope at 2.0 nm/pixel (xy). We imaged only the region containing the aboral organ. Subsequently, images from 620 layers were stitched and aligned in TrakEM2 [@cardona2012]. The final vEM dataset encompassed a volume of 60 μm × 40 μm × 30 μm. We skeletonised and annotated all cells in this dataset in CATMAID [@Saalfeld2009; @Schneider-Mizell2016], resulting in a reconstruction of 972 cells that had intact nuclei and cell bodies, 899 of which formed the aboral organ (AO). Most of the 899 AO cells, including the balancers and the aboral nerve net neurons are fully contained within the volume, due to the compact and mostly self-contained organisation of the aboral organ.

In cells containing cilia, we also traced cilia along their length and annotated all basal bodies. For neuronal skeletonization, nodes were placed to interconnect the profiles of the same neuron’s processes across layers, extending the skeleton until all branches were fully traced. All nodes relevant to synaptic sites were tagged, and skeletons were named and assigned multi-level annotations. As described later, some neurons formed loop-like structures (anastomosed neurons), wherein separated branches often rejoined either the main branch or smaller branches [@Burkhardt2023]. In such cases, branch nodes were placed near the closest existing node and annotated accordingly (CATMAID only supports skeleton trees). The entire skeletonized volume was composed of 134,591 nodes. 73 fragments could not be attached to somata-associated skeletons. Most of these fragments represent short skeletal branches that could not be traced beyond gaps or low-quality layers.

@@

Classical neural staining techniques did not provide clear images of neurons at the aboral organ. However, ultrastructural studies revealed morphological evidence for neural elements, including synapses that are located on the epithelial floor of the aboral organ—a region where epithelial cells are densely packed and form a floor-like structure [@Horridge_1964; @Hernandez_Nicaise1973a]. In our dataset, we identified 428 synapses based on the previously described pre-synaptic triad morphology consisting of a single layer of vesicles, a cisterna of smooth endoplasmic reticulum and an associated mitochondrion. At synaptic sites, we marked mitochondria as synaptic nodes (orange in CATMAID) and connected this node to the nearest node in postsynaptic cells across the regions where synaptic vesicles aligned with the presynaptic cell's membrane (light blue arrows in CATMAID) (Figure 2--figure supplement 1A). We could not detect specialized post-synaptic structures, in agreement with previous studies [@Hernandez_Nicaise1973a]. Synapses were identified as either monoadic or polyadic, with one pre-synaptic neuron forming connections with one or multiple post-synaptic cells (Figure 2--figure supplement 1B).

We reconstructed three aboral nerve-net (ANN) neurons, each with a syncytial morphology (Figure 2A and B). Each ANN had multiple nuclei (ranging from two to five per neuron), with anastomosing membranes (Supplementary Figure S2C). These neurons are distinct from the syncytial subepithelial nerve net (SNN) neurons with blebbed morphology previously identified by vEM in the body wall of *Mnemiopsis* [@Burkhardt2023]. Our results reveal a second type of syncytial neurons in ctenophores (Supplementary Figure S2D).

Furthermore, ANNs are also different from other neurons previously reported by EM, including mesogleal neurons and ciliated sensory cells that synapse on the SNN (types 1-4)[@Burkhardt2023]. Based on their distinct size and position, we classified the ANNs into two types. The first type is a single large neuron (ANN_Q1-4) that spans all four quadrants (Q1-Q4) and has four (possibly five) nuclei. This neuron has 150 presynaptic sites. The second type consists of two neurons (ANN_Q1Q2 and ANN_Q3Q4), each spanning two quadrants and containing two nuclei. ANN_Q1Q2 had 119 presynaptic structures, whereas ANN_Q3Q4 had 68 presynaptic structures.

@@

To investigate the regulation of balancer function, we carried out a high-speed video-microscope analysis of balancer cilia. Previous studies [@Tamm1980; @Tamm1982; @Lowe1997] have established that balancer cilia function as mechanoreceptors, with their beating frequency modulated by inclination. Tamm also suggested that differences in statolith morphology and the shape of balancer cilia between the tentacular and sagittal planes could lead to different forces exerted on cilia by the statolith [@Tamm_2014c; @tamm2015]. To further explore this, we used a tilted microscope with a vertical stage where we mounted immobilised cydippid larvae with their aboral-oral axis aligned in different orientations relative to the gravity vector. We then compared larvae oriented in different angles and either with their sagittal or tentacular plane parallel to the sample stage.

@@

![**Figure 2—figure supplement 1. Ultrastructural features and synaptic organization of the ANN, with comparison to the subepithelial nerve net (SNN)**\
(A) Registration and annotation of synapses in the actual cellular traces. Pink dots indicate the approximate center of each individual cell, marked within its plasma membrane boundary. Mitochondria involved in synaptic contacts are labeled with orange dots. Red arrows point from these mitochondria to the presynaptic cell body, marking presynaptic structures. Cyan arrows point from the orange-labeled mitochondria to the corresponding postsynaptic cell.\
(B) Electron microscopy (EM) images showing representative presynaptic structures. The left panel shows a monadic synapse, while the right panel shows a polyadic (byadic) synapse, in which a single presynaptic site contacts two postsynaptic cells.\
(C) Nuclei within two different types of ANN cells are shown in purple in a single EM section. On the left, the ANN_Q1-4 cell extends widely across the aboral organ and appears as a continuous structure in 3D, despite having nuclei located distantly from each other. In contrast, ANN_Q1Q2 (right) contains two adjacent nuclei situated close together, suggesting a more localized distribution.\
(D) Three-dimensional reconstruction of the subepithelial nerve net (SNN). The SNN extends broadly, enveloping the aboral organ. Some portions of the network remain as unconnected fragments due to incomplete reconstruction.\
(E) Morphological comparison between the ANN in EM images. The SNN shows a "beads-on-a-string" morphology, with discrete bead-like swellings connected by faint electron-dense fibrous structures. \
(F) Three-dimensional reconstruction of the SNN and ANNs.\
(G) Morphological comparison between the SNN in EM images. The ANN exhibits a more diffuse distribution, often appearing to interdigitate between surrounding cells. Relatively large mitochondria are frequently observed in the ANN processes.](figures/Figure2_Supplement1.png)

>>>>

For vEM analysis, we used a five-day-old cydippid larva of *Mnemiopsis leidyi*. To obtain high-quality ultrastructural preservation, we used high pressure freezing followed by freeze substitution and resin embedding (Epon). The specimen was sectioned from the aboral tip and we collected approximately 1,000 ultra-thin (50 nm) serial sections as ribbons. The sections were imaged in a Zeiss Gemini 500 scanning electron microscope at 2.0 nm/pixel (xy). We imaged only the region containing the aboral organ. Subsequently, images from 620 layers were stitched and aligned in TrakEM2 [@cardona2012]. The final vEM dataset encompassed a volume of 60 μm × 40 μm × 30 μm. We skeletonised and annotated all cells in this dataset in CATMAID [@Saalfeld2009; @Schneider-Mizell2016], resulting in a reconstruction of 972 cells that had intact nuclei and cell bodies, 898 of which formed the aboral organ (AO). Most of the 898 AO cells, including the balancers and the aboral nerve net neurons are fully contained within the volume, due to the compact and mostly self-contained organisation of the aboral organ.

In cells containing cilia, we also traced cilia along their length and annotated all basal bodies (Video 2. and not shown). For neuronal skeletonization, nodes were placed to interconnect the profiles of the same neuron’s processes across layers, extending the skeleton until all branches were fully traced. All nodes relevant to synaptic sites were tagged, and skeletons were named and assigned multi-level annotations. As described later, some neurons formed loop-like structures (anastomosed neurons), wherein separated branches often rejoined either the main branch or smaller branches [@Burkhardt2023]. In such cases, branch nodes were placed near the closest existing node and annotated accordingly (CATMAID only supports skeleton trees). The entire skeletonized volume was composed of 134,591 nodes. 73 fragments could not be attached to somata-associated skeletons. Most of these fragments represent short skeletal branches that could not be traced beyond gaps or low-quality layers.

@@

Immunofuorescent stainings in *Pleurobrachia pileus* revealed a nerve net at the base of the aboral organ that seemed continuous with the polygonal nerve net of the surrounding epithelium [@Jager2011]. Ultrastructural studies revealed synapses near the floor of the aboral organ—a region where epithelial cells are densely packed and form a floor-like structure [@Horridge_1964; @Hernandez_Nicaise1973a]. In our dataset, we identified 428 synapses based on the previously described pre-synaptic triad morphology consisting of a single layer of vesicles, a cisterna of smooth endoplasmic reticulum and an associated mitochondrion. At synaptic sites, we marked mitochondria as synaptic nodes (orange in CATMAID) and connected this node to the nearest node in postsynaptic cells across the regions where synaptic vesicles aligned with the presynaptic cell's membrane (light blue arrows in CATMAID) (Figure 2--figure supplement 1A). We could not detect specialized post-synaptic structures, in agreement with previous studies [@Hernandez_Nicaise1973a]. Synapses were identified as either monoadic or polyadic, with one pre-synaptic neuron forming connections with one or multiple post-synaptic cells (Figure 2--figure supplement 1B).

We reconstructed three aboral nerve-net (ANN) neurons, each with a syncytial morphology (Figure 2A and B). Each ANN had multiple nuclei (ranging from two to five per neuron), with anastomosing membranes (Supplementary Figure S2C). These neurons are distinct from the syncytial subepithelial nerve net (SNN) neurons with blebbed morphology previously identified by vEM in the body wall of *Mnemiopsis* [@Burkhardt2023]. ANN neurons also lack dense core vesicles, structures that are abundant throughout the cell bodies and neurites of SNN neurons [@Burkhardt2023]. Our results reveal a second type of syncytial neurons in ctenophores (Figure 2--figure supplement C-G).

Furthermore, ANNs are also different from other neurons previously reported by EM, including mesogleal neurons and ciliated sensory cells that synapse on the SNN (types 1-4)[@Burkhardt2023]. Based on their distinct size and position, we classified the ANNs into two types. The first type is a single large neuron (ANN_Q1-4) that spans all four quadrants (Q1-Q4) and has four (possibly five) nuclei. This neuron has 150 presynaptic sites (Figure 2C-F). The second type consists of two neurons (ANN_Q1Q2 and ANN_Q3Q4), each spanning two quadrants and containing two nuclei. ANN_Q1Q2 had 119 presynapses, whereas ANN_Q3Q4 had 68 presynaptic structures (Figure 2C-F).

@@

To investigate the regulation of balancer function, we carried out a high-speed video-microscope analysis of balancer cilia. Previous studies [@Tamm1980; @Tamm1982; @Lowe1997] have established that balancer cilia function as mechanoreceptors, with their beating frequency modulated by inclination. Tamm also suggested that differences in statolith morphology and the shape of balancer cilia between the tentacular and sagittal planes could lead to different forces exerted on cilia by the statolith [@Tamm_2014c; @tamm2015]. To further explore this, we used a tilted microscope with a vertical stage where we mounted immobilised cydippid larvae with their aboral-oral axis aligned in different orientations relative to the gravity vector. We then compared larvae oriented in different angles and either with their sagittal or tentacular plane parallel to the sample stage (Figure 4A).

@@

![**Figure 2—figure supplement 1. Ultrastructural features and synaptic organization of the ANN, in comparison to the subepithelial nerve net (SNN)**\
(A) Registration and annotation of synapses in the actual cellular traces. Pink dots indicate the approximate center of each individual cell, marked within its plasma membrane boundary. Mitochondria involved in synaptic contacts are labeled with orange dots. Red arrows point from these mitochondria to the presynaptic cell body, marking presynaptic structures. Cyan arrows point from the orange-labeled mitochondria to the corresponding postsynaptic cell.\
(B) Electron microscopy (EM) images showing representative presynaptic structures. The left panel shows a monadic synapse, while the right panel shows a polyadic (byadic) synapse, in which a single presynaptic site contacts two postsynaptic cells.\
(C) Nuclei within two different types of ANN cells are shown in purple in a single EM section. On the left, the ANN_Q1-4 cell extends widely across the aboral organ and appears as a continuous structure in 3D, despite having nuclei located distantly from each other. In contrast, ANN_Q1Q2 (right) has two adjacent nuclei.\
(D) Three-dimensional reconstruction of the subepithelial nerve net (SNN). The SNN extends broadly, enveloping the aboral organ. Some portions of the network remain as unconnected fragments due to incomplete reconstruction.\
(E) Morphological comparison between the ANN in EM images. The SNN shows a "beads-on-a-string" morphology, with discrete bead-like swellings connected by faint electron-dense fibrous structures. \
(F) Three-dimensional reconstruction of the SNN and ANNs.\
(G) Morphological comparison between the SNN in EM images. The ANN exhibits a more diffuse distribution, often appearing to interdigitate between surrounding cells. Relatively large mitochondria are frequently observed in the ANN processes.](figures/Figure2_Supplement1.png)
